Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities

### 2021-10-12 paper/manuscript.Rmd
Add clarification of what kinds of traits we used and why.

---
title: "Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found no overall trend in species diversity through time accompanied by increasing turnover. Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait databases. We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover. This pattern held at the study-level where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric. Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental change. We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.

@@

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
In order to ensure comparability across taxa in trait type and data quality we therefore focus on these taxa.

@@

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxon see Table 1.

>>>>

---
title: "Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Perry de Valpine"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found no overall trend in species diversity through time accompanied by increasing turnover. Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait representitive of a species' functional role. We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover. This pattern held at the study-level where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric. Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental and community change. We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.

@@

Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected. In order to ensure comparability across taxa in trait type and data quality we therefore focus on these taxa. Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].

@@

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that best represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxon see Table 1.
